---
title: "Gym session analysis"
author: "Jean-Francois Cantin"
date: "April 22, 2015"
output: html_document
---

```{r initialization}
rm(list=ls())
library(tidyr)
library(dplyr)
library(httr)
#library(data.table)
library(ggplot2)
library(lubridate)
library(caret)
library(rattle)
```

```{r setupCluster}
library(doParallel)
detectCores()
mc <- 4
cl <- makeCluster(mc)
registerDoParallel(cl)
```


# Submission
**The goal of your project is to predict the manner in which they did the exercise. This is the "classe" variable in the training set. You may use any of the other variables to predict with. You should create a report describing how you built your model, how you used cross validation, what you think the expected out of sample error is, and why you made the choices you did. You will also use your prediction model to predict 20 different test cases. **

1. Your submission should consist of a link to a Github repo with your R markdown and compiled HTML file describing your analysis. Please constrain the text of the writeup to < 2000 words and the number of figures to be less than 5. It will make it easier for the graders if you submit a repo with a gh-pages branch so the HTML page can be viewed online (and you always want to make it easy on graders :-).
2. You should also apply your machine learning algorithm to the 20 test cases available in the test data above. Please submit your predictions in appropriate format to the programming assignment for automated grading. See the programming assignment for additional details. 

# Classes
Six young health participants were asked to perform one set of 10 repetitions of the Unilateral Dumbbell Biceps Curl in five different fashions: 

* exactly according to the specification (Class A) 
* throwing the elbows to the front (Class B)
* lifting the dumbbell only halfway (Class C)
* lowering the dumbbell only halfway (Class D) 
* throwing the hips to the front (Class E)

Read more: http://groupware.les.inf.puc-rio.br/har#ixzz3Y9JZ39Ox

## TODO:
* Should I try heatmap.2?

# Get Data
```{r originalData}
# The data for this project come from this source: http://groupware.les.inf.puc-rio.br/har
trainingFile <- "data//training.rds"
testingFile <- "data//testing.rds"
if(file.exists(trainingFile) && file.exists(testingFile)){
  otraining <- readRDS(trainingFile)
  otesting <- readRDS(testingFile)
} else{
  pmlTrainingUrl <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
  pmlTestingUrl <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"
  otraining <- tbl_df(content(GET(pmlTrainingUrl)))
  otesting <- tbl_df(content(GET(pmlTestingUrl)))
  saveRDS(otraining, trainingFile)
  saveRDS(otesting, testingFile)
}
```

# Data Exploration

## Data Cleaning
```{r initialDataReview, warning=FALSE}
dim(otraining)
glimpse(otraining)
```
* Lots of columns with NA and empty strings
* Some of the `character` columns have been misclassified. They should be double.

```{r columnClassification}
# reorder columns
training <- otraining %>% select(classe, user_name:magnet_forearm_z)

# create factors
factorCols <- c(1,2,6)
training[factorCols] <- lapply(training[factorCols], as.factor)

# create dates
training$cvtd_timestamp <- dmy_hm(training$cvtd_timestamp)

# convert char to double
ischarCols <- sapply(training, is.character)
training[ischarCols] <- lapply(training[ischarCols], as.double)

# combine timestamps
training <- training %>% 
  mutate(rawtimestamp = as.double(paste(raw_timestamp_part_1, raw_timestamp_part_2, sep="."))) %>% 
  select(rawtimestamp, classe, user_name, roll_belt:magnet_forearm_z)

```

There appears to be lots of NA's so we need a strategy in dealing with them. 
After reading the paper associated with the data we can conclude that most of the variables are calculated based on their window strategy. With that in mind the only raw data we should use are the original time series since any derived / calculated features will not be present in the test dataset. :
* all gyros,accel,magnet xyz
* none of the kurtosis, skewness, max, min, amplitude, var, avg, stddev, 

```{r columnSelection}
training <- training %>% 
  select(-starts_with("kurtosis_"),
         -starts_with("skewness_"),
         -starts_with("max_"),
         -starts_with("min_"),
         -starts_with("amplitude_"),
         -starts_with("var_"),
         -starts_with("avg_"),
         -starts_with("stddev_"))
```


```{r explore}
str(training$classe)
p <- ggplot(training %>% filter(user_name=="carlitos", classe=="A"))
p <- p + geom_line(aes(x=rawtimestamp, y=total_accel_dumbbell)) + aes(colour=classe)
p <- p + facet_grid(.~classe, scales = "free_x")
p

p <- ggplot(training)# %>% filter(user_name=="carlitos"))
p <- p + geom_line(aes(x=rawtimestamp, y=total_accel_dumbbell)) + aes(colour=classe)
p <- p + facet_grid(classe ~ user_name, scales = "free_x")
p

```

# Classification Preparation
Let's start with a small subset of the data like one user to test protocol

### Need to try different model
```{r sample, eval=FALSE}
carlitos <- training %>% filter(user_name == "carlitos")
inTrain <- createDataPartition(carlitos$classe, p = 3/4)[[1]]
carTrain <- carlitos[inTrain,]
carTest <- carlitos[-inTrain,]
carlitosPCA <- preProcess(carTrain %>% select(-rawtimestamp, -classe, -user_name), method="pca", thresh = .9)
summary(carlitosPCA)
carlitosPcaPredict <- predict(carlitosPCA, carTrain %>% select(-rawtimestamp, -classe, -user_name))
carlitosModel <- train(carTrain$classe ~ ., method="rf", data=carlitosPcaPredict)
carlitosTestPCA <- predict(carlitosPCA, carTest %>% select(-rawtimestamp, -classe, -user_name))
mod1conf <- confusionMatrix(carTest$classe, predict(carlitosModel, carlitosTestPCA))

carlitosModel2 <- train(classe ~ ., method="rpart", data=carTrain[,-1])
carlitosModel2
confusionMatrix(carTest$classe, predict(carlitosModel2, carTest[,-1]))

carlitosModel3 <- train(classe ~ ., method="rf", data=carTrain[,-1])
carlitosModel3
mod3conf <- confusionMatrix(carTest$classe, predict(carlitosModel3, carTest[,-1]))
mod3conf$table

mod1conf$overall
- 
  mod3conf$overall

```

## Full Classification
```{r fullClassification}

set.seed(123)
inTrain <- createDataPartition(training$classe, p = .8)[[1]]
# remove timestamp
trainset <- training[inTrain,-1]
testset <- training[-inTrain,-1]

transparentTheme(trans = 0.9)
featurePlot(x = trainset[,3:20],
            y = trainset$classe,
            plot = "density",
            scales = list(y = list(relation="free"),
                          x = list(relation="free")),
            adjust = 1.5,
            pch = "|",
            layout = c(4,5),
            auto.key = list(columns = 5))

pre <- preProcess(trainset %>% select(-classe, -user_name), method="pca")
pre
preTrain <- predict(pre, trainset %>% select(-classe, -user_name))
set.seed(456)
system.time(modFitPca <- train(trainset$classe ~ ., method="rf", data=preTrain))
set.seed(456)
system.time(modFit <- train(classe ~ ., method="rf", data=trainset))

preTest <- predict(pre, testset %>% select(-classe, -user_name))

fancyRpartPlot(modFitPca$finalModel)
modPcaCM <- confusionMatrix(testset$classe, predict(modFitPca, preTest))
modCM <- confusionMatrix(testset$classe, predict(modFit, testset))

modPcaCM
modCM
plot(modFit, log = "y", main="Model Accuracy" )
```

```{r createFinalPrediction}
# Prepare TestFile
colnames(trainset)
finalTest <- otesting[colnames(trainset)]
finalPrediction <- predict(modFit, finalTest)
pml_write_files(as.character(finalPrediction))
```


```{r LastWords}
stopCluster(cl)
sessionInfo()
```

